Ecosystem size and complexity as extrinsic drivers of food chain length in branching ecosystems

### 2023-07-10 rmd/manuscript.Rmd
edit: manuscript typo

$m_{1,up}$ and $m_{2,up}$ are the disturbance values of upstream contributing tributaries, $\varepsilon_1$ and $\varepsilon_2$ are independent local noises, and $\omega$ the relative size of one tributary over another defined as $\omega = \frac{Q_{1}}{Q_{1} + Q_{2}}$. Thus, the larger tributary has a greater influence on the downstream environment, as observed in natural river networks [@bendaNetworkDynamicsHypothesis2004]. Disturbance value distributions had medians close to $\mu_m$ with long tails (**Figure 1B**). The low-magnitude patches can be viewed as refugia, which supply individuals to impacted sites in subsequent time steps.

@@

where the parameter $s$ determines the strength of switching ($0 \le s \le 1$). In this expression, the intraguild predator increases its attack rate to the basal species if $\phi > 0$, properly representing the switching behavior of the intraguild prey.

@@

We also calculated the proportion of the patch state ("No species", "B", "B + C", "B + P", and "B + C + P") at each time step to infer mechanisms that drive changes in FCL. The time-averaged proportion of patch state $\bar{\psi}$ was obtained as $\bar{\psi} = q_t/(N_p T)$, where $q_t$ is the number of patches in a given patch state at time step $t$.

@@

Under each motif, we crossed values of disturbance (frequency and intensity), productivity, and dispersal. Disturbance frequency and intensity were controlled via $p_m$ ( `r unique(df_sim$p_disturb) %>% sort()`) and $\mu_m$ (`r unique(df_sim$mean_disturb_source) %>% sort()`), respectively. Spatial disturbance patterns were regulated by variation in disturbance intensity at headwaters ($\sigma_{h}=$ `r max(df_sim$sd_disturb_source)`), and local variation ($\sigma_{l}=$ `r min(df_sim$sd_disturb_lon)`). This setup ($\sigma_h > \sigma_l$) produces high spatial variation among tributaries while allowing strong within-tributary downstream correlation [@teruiEmergentDualScaling2021], a typical pattern in real river networks [@nakamuraDisturbanceRegimesStream2000; @bendaNetworkDynamicsHypothesis2004]. We used a combination of basal species' growth rate $r_b$ and base carrying capacity $K_{base}$ (low productivity: $r_b =$ `r min(df_sim$r_b)`, $K_{base} =$ `r min(df_sim$base_k)` ; high productivity = $r_b =$ `r max(df_sim$r_b)`, $K_{base} =$ `r max(df_sim$base_k)` ) as a proxy for resource availability. For dispersal, we examined short- and long-distance dispersal scenarios (short-distance dispersal $\theta=$ `r max(df_sim$theta)`; long-distance dispersal $\theta=$ `r min(df_sim$theta)`) with constant dispersal probability ($p_{d}=$ `r unique(df_sim$p_dispersal)`). Parameter values were summarized in **Table 1**.

@@

With no disturbance, the effects of ecosystem geometry were highly context dependent (**Figure 2**). In the chain scenario (no omnivory) with low productivity, ecosystem size had a positive effect on FCL (**Figure 3**), while ecosystem complexity had a negative influence (**Figure 4**). These opposing influences of ecosystem size and complexity were associated with their contrasting relationships with the total carrying capacity (i.e., carrying capacity summed across all patches): ecosystem size increased the total carrying capacity, whereas ecosystem complexity decreased it (**Appendix S1: Supplementary Text8**). However, when considering some levels of omnivory, these patterns characteristic of the chain scenario diminished. With weak omnivory (the second rows of **Figure 3 and 4**), ecosystem complexity had consistent positive effects on FCL, except the scenario with low productivity and short dispersal (**Figure 4**). In contrast, the effect of ecosystem size was more productivity dependent (**Figure 4**). Neither of ecosystem size and complexity influenced FCL with strong omnivory.

@@

In the weak and strong omnivory motifs, different state transitions were associated with ecosystem size and complexity. Ecosystem size increased FCL primarily through additive mechanisms, facilitating the state transition of "B" $\rightarrow$ "B + P" (low productivity) or "No species" $\rightarrow$ "B + P" (high productivity), particularly with frequent disturbance (**Figure 5**). In contrast, insertion (i.e., the insertion of intermediate trophic species) and additive mechanisms were responsible for the effect of ecosystem complexity. With weak omnivory, the transition from "B + P" to "B + C + P" (insertion) was predominant when productivity was low (**Figure 6B**). However, the additive mechanism ("No species" $\rightarrow$ "B + P") took primacy in the high productivity scenario (**Figure 6B**). In the strong omnivory motif, transition patterns depended on disturbance frequency. When disturbance was rare ($p_m = 0.05$), the insertion mechanism ("B + P" $\rightarrow$ "B + C + P") drove the FCL change (**Figure 6C**). As disturbance frequency increased, the primary process switched to an additive mechanism ("B" $\rightarrow$ "B + P" or "No species" $\rightarrow$ "B + P"; **Figure 6C**).

>>>>

$m_{1,up}$ and $m_{2,up}$ are the disturbance values of upstream contributing tributaries, $\varepsilon_1$ and $\varepsilon_2$ are independent local noises, and $\omega$ the relative size of one tributary over another defined as $\omega = Q_{1}/(Q_{1} + Q_{2})$. Thus, the larger tributary has a greater influence on the downstream environment, as observed in natural river networks [@bendaNetworkDynamicsHypothesis2004]. Disturbance value distributions had medians close to $\mu_m$ with long tails (**Figure 1B**). The low-magnitude patches can be viewed as refugia, which supply individuals to impacted sites in subsequent time steps.

@@

where the parameter $s$ determines the strength of switching ($0 \le s \le 1$). In this expression, the intraguild predator increases its attack rate to the basal species if $\phi > 0$, properly representing the switching behavior.

@@

We also calculated the proportion of the patch state ("No species", "B", "B + C", "B + P", and "B + C + P") at each time step to infer mechanisms that drive changes in FCL. The time-averaged proportion of patch state $\bar{\psi}$ was obtained as $\bar{\psi} = \sum_t q_t/(N_p T)$, where $q_t$ is the number of patches in a given patch state at time step $t$.

@@

Under each motif, we crossed values of disturbance (frequency and intensity), productivity, and dispersal. Disturbance frequency and intensity were controlled via $p_m$ ( `r unique(df_sim$p_disturb) %>% sort()`) and $\mu_m$ (`r unique(df_sim$mean_disturb_source) %>% sort()`), respectively. Spatial disturbance patterns were regulated by variation in disturbance intensity at headwaters ($\sigma_{h}=$ `r max(df_sim$sd_disturb_source)`), and local variation ($\sigma_{l}=$ `r min(df_sim$sd_disturb_lon)`). This setup ($\sigma_h > \sigma_l$) produces high spatial variation among tributaries while allowing strong within-tributary downstream correlation [@teruiEmergentDualScaling2021], a typical pattern in real river networks [@nakamuraDisturbanceRegimesStream2000; @bendaNetworkDynamicsHypothesis2004]. We used a combination of basal species' growth rate $r_b$ and base carrying capacity $K_{base}$ (low productivity: $r_b =$ `r min(df_sim$r_b)`, $K_{base} =$ `r min(df_sim$base_k)` ; high productivity: $r_b =$ `r max(df_sim$r_b)`, $K_{base} =$ `r max(df_sim$base_k)` ) as a proxy for resource availability. For dispersal, we examined short- and long-distance dispersal scenarios (short-distance dispersal $\theta=$ `r max(df_sim$theta)`; long-distance dispersal $\theta=$ `r min(df_sim$theta)`) with constant dispersal probability ($p_{d}=$ `r unique(df_sim$p_dispersal)`). Parameter values were summarized in **Table 1**.

@@

With no disturbance, the effects of ecosystem geometry were highly context dependent (**Figure 2**). In the chain scenario (no omnivory) with low productivity, ecosystem size had a positive effect on FCL (**Figure 3**), while ecosystem complexity had a negative influence (**Figure 4**). These opposing influences of ecosystem size and complexity were associated with their contrasting relationships with the total carrying capacity (i.e., carrying capacity summed across all patches): ecosystem size increased the total carrying capacity, whereas ecosystem complexity decreased it (**Appendix S1: Figure S8**). However, when considering some levels of omnivory, these patterns characteristic of the chain scenario diminished. With weak omnivory (the second rows of **Figure 3 and 4**), ecosystem complexity had consistent positive effects on FCL, except the scenario with low productivity and short dispersal (**Figure 4**). In contrast, the effect of ecosystem size was more productivity dependent (**Figure 4**). Neither of ecosystem size and complexity influenced FCL with strong omnivory.

@@

In the weak and strong omnivory motifs, different state transitions were associated with ecosystem size and complexity. Ecosystem size increased FCL primarily through additive mechanisms, facilitating the state transition of "B" $\rightarrow$ "B + P" (low productivity) or "No species" $\rightarrow$ "B + P" (high productivity), particularly with frequent disturbance (**Figure 6--7**). In contrast, insertion (i.e., the insertion of intermediate trophic species) and additive mechanisms were responsible for the effect of ecosystem complexity. With weak omnivory, the transition from "B + P" to "B + C + P" (insertion) was predominant when productivity was low (**Figure 9**). However, the additive mechanism ("No species" $\rightarrow$ "B + P") took primacy in the high productivity scenario (**Figure 9**). In the strong omnivory motif, transition patterns depended on disturbance frequency. When disturbance was rare ($p_m = 0.05$), the insertion mechanism ("B + P" $\rightarrow$ "B + C + P") drove the FCL change (**Figure 10**). As disturbance frequency increased, the primary process switched to an additive mechanism ("B" $\rightarrow$ "B + P" or "No species" $\rightarrow$ "B + P"; **Figure 10**).
